A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-10 main.tex
Use the term of multi-stage.

\blue{181 words. The abstract should briefly summarize the contents of the paper in 150--250 words.}
\blue{コードとかはボールドを使ったほうが良いかもしれない。}
\red{Suspicious senetences are colored red.}
%
%
%
\red{Multistage Programming と Multistage Calculus をどう使い分けるか?}

@@

\subsection{Multistage Programming}

@@

Multistage programming enables us to generate and run fragments of code at runtime.
MetaOCaml\cite{oleg2014} provide features for multistage programming, which are brackets, escape, and run.
Brackets, written \verb|.< e >.| in MetaOCaml, makes a code value from \verb|e|.
Run, written \verb| run e | in MetaOCaml, runs a code value of \verb|e| and restore \verb|e|.
Escape, written \verb| .~ e | in MetaOCaml, expands a code value of \verb|e|.
Unlike run, escape is supposed to use in a fragment of code, that is, escape cannot be used as an alternative of run.
For example, the following MetaOCaml expression

@@

Cross-stage Persistence (CSP) is another primitive of multistage programming, which enable us to embed computed values into a code value.
This is an example of CSP.
\begin{verbatim}
        let plusone = fun x -> x + 1 in
        let y = plusone 2 in  .< y >.
\end{verbatim}
This program evaluates to \verb|.< 3 >.| not  \verb|.< plusone 2 >.|.
This is because the variable \verb|y| is introduced into a code fragment using CSP.
Therefore, the variable \verb|y| is embedded after it was calculated.
CSP is very important when we write practical programs 
because CSP enables us to use library functions in code fragments as in \verb|.<List.combine [1;2] ['a';'b']>.|.

@@

The main application of multistage calculi is program optimization.
A famous example is power functions.
Usual power functions take two arguments, which are the base and the exponent.
We can make a specialized power function for given exponents with multistage programming and 
optimize them by unrolling a loop in functions for given exponents.
The optimized power function for the exponent of \verb|3| looks like \verb|power3 = fun x -> x * x * x|.
\verb|power3| is faster than ordinary \verb|power| function because it contains no loop.
Multistage programming can optimize functions which take more than two arguments 
by generating a code fragment optimized to a given argument.

@@

Multistage programming can also optimize vector calculation.
For example, \verb|vadd| function, which takes two vectors and return the sum of them,
is implemented with a loop in many cases.
We can unroll \verb|vadd| function for a given vector length and optimize it with multistage programming.
\red{MetaOCamlで書いた長さ付きベクトルの例?}

@@

However, there is a sever problem in functions which are optimized by multistage programming.
Although unrolled \verb|vadd| function is optimized, we cannot use it for different vector length.
For example, when you optimize for the length of 5, you shouldn't use it for vectors of 3 lengths.
Otherwise, we will get a Segmentation Fault error.
This problem is serious but existing type systems for multistage calculi cannot prevent it.

@@

\subsection{Multistage Programming with Dependent Types}
\red{このsubsectionは短いので上とくっつけてもいいかもしれない}

@@

As we pointed out in the above section,
functions optimized with multistage programming can take only restricted values as arguments.
When you optimize \verb|vadd| function for the length of 5, you should it only for 5 length vectors.
We introduce dependent types into a multistage calculus
so that the type system can guarantee optimized functions are used properly.

In this paper, we design new multistage calculus \LMD by 
merging dependent types into existing multistage calculus \LTP\cite{Hanada2014}.
\LMD is multistage calculus which contains dependent types.
\LMD makes multistage programming more safer with the power of dependent types.

@@

We designed \LMD, a multistage calculus with dependent type.
\LMD is based on \LTP\cite{Hanada2014} by Hanada and Igarashi, which is a multistage calculus with CSP and
we introduced dependent types based on \LTP\cite{attapl}.
In this section, we check \LMD informally after checking \LTP and \LLF which are the basis of \LMD.

@@

\LTP\cite{Hanada2014} is a multistage calculus with CSP by Hanada and Igarashi.
In \LTP, brackets and escape are written $\TB_\alpha M$ and $\TBL_\alpha M$, respectively.
The type of $\TB_\alpha M$ is $\TW_\alpha \tau$ if $M$ has type $\tau$.
The type of $\TBL_\alpha M$ is $\tau$ if $M$ has $\TW_\alpha \tau$.
% Please notice that if $\TBL_\alpha M$ is well-typed, $M$ has a code type from the typing rule of \LTP.
In addition to normal $\beta$-reduction, there is a reduction rule for brackets and escape.
\begin{align*}
    \TBL_\alpha (\TB_\alpha M) \longrightarrow M
\end{align*}
It means escape cancels brackets.
% This reduction is called $\longrightarrow_\Lambda$ in \LMD.
% \red{この文はここに来るべきなのか? LTPの説明ではないが}

@@

Another important rule about a transition variable is 
that symbols with transition variables disappear 
when the empty transition $\epsilon$ is substituted to the transition variable.
The purpose of this rule is express \verb|run| without any special symbol.
In multistage calculus, \verb|run| is a very important operator which changes quoted code to the original code.
For example, $(\TB_\epsilon (\lambda x:\I.x))$ is equivalent to $(\lambda x:\I.x)$.
Therefore, $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))\ \epsilon$ becomes $\lambda x:\I.x$
In \LTP, \verb|run| is realized with application to the empty transition $\epsilon$.

@@

Second, we need kinding rules and well-formed kinding rules in order to extend \LMD with dependent types.
It was lucky that almost all rules are determined easily.
This is because multistage calculus and dependent types are almost orthogonal.
\red{orthogonal は抽象的すぎるか?}
Therefore, we can get kinding rules and well-formed kinding rules of \LMD just by 
attaching stage anotations to ones of \LLF.
For example, \KAbs-LF is a kinding rule for a dependent type in \LLF and \KAbs\ is a corresponding one.
\begin{center}
    \infrule{\G\vdash \tau :: * \andalso \G,x:\tau @A\vdash \sigma::J}{\G\vdash(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J)}{\KAbs-LF} \\[2mm]
    \infrule{\G\V \tau :: *@A \andalso \G,x:\tau @A\V \sigma::J@A}{\G\V(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J)@A}{\KAbs} \\[2mm]
\end{center}

@@

% 多段階計算のルール
The rules \TTB, \TTBL, \TGen, \TIns, and \TCsp\ are rules for a multistage calculus.
They are corresponding to quoting a code, unquoting a code, making a stage abstraction, 
application of $\epsilon$ stage to a stage abstraction, and cross-stage persistence, respectively.
Please check Hanada and Igarashi \cite{Hanada2014} and Tsukada and Igarashi \cite{Tsukada} for details of these rules.

@@

% \QPercentの説明
The rule \QPercent is something special because it has no background of syntax or reductions.
But we add this rule to \LMD\ because it is essential when you use dependent types in code type.
For example, we can utilize a dependently typed multistage calculus to generate code for matrix multiplication.
When we know sizes of matrices before the multiplication,
we can generate efficient code for multiplication by unrolling loops.

@@

\red{Besides,} multistage calculi also have a long history of research.
Davis revealed there is Curry-Howard correspondence between a multistage calculus and modal logic in $\lambda^\circ$\cite{davies1996temporal}.
However, $\lambda^\circ$ don't have operators for run and CSP.
\red{runはboldか? Besidesより適切な接続詞はないのか?}
Benaissa, et al. \cite{benaissa1999logical} study the relationship between multistage calculus and category theory or modal logic.
Taha and Sheard introduced run and CSP to a multistage calculus in \cite{MetaML}.
Additionally, Taha and Nielsen invented the concept of environment classifiers in $\lambda^\alpha$\cite{taha2003environment} and 
construct the type system for $\lambda^\alpha$, which can handle quoting, unquoting, run, and CSP.

In $\lambda^\TW$\cite{Tsukada}, Tsukada and Igarashi found modal logic which corresponds to a multistage calculus with environment classifiers and
show that run can be represented as application of $\epsilon$ to transition abstractions.
In \LTP\cite{Hanada2014}, Hanada and Igarashi extended $\lambda^\TW$ with CSP and discuss code residualization 
which allows us to dump the quoted code into an external file.

@@

Combination of multistage programming and another programming feature has been studied.
Oleg, Kameyama, and Sudo\cite{kiselyov2016refined} designed a type sysytem for multistage programming with mutable cells and
Oishi and Kameyama\cite{oishi2017staging} extended a multistage calculus so that it can handle control operators such as shift and reset.

>>>>

\blue{181 words. The abstract should briefly summarize the contents of the paper in 150--250 words.}
\blue{コードとかはボールドを使ったほうが良いかもしれない。}
\red{Suspicious senetences are colored red.}
%
%
%
\red{Multi-stage Programming と Multi-stage Calculus をどう使い分けるか?}

@@

\subsection{Multi-stage Programming}

@@

Multi-stage programming enables us to generate and run fragments of code at runtime.
MetaOCaml\cite{oleg2014} provide features for multi-stage programming, which are brackets, escape, and run.
Brackets, written \verb|.< e >.| in MetaOCaml, makes a code value from \verb|e|.
Run, written \verb| run e | in MetaOCaml, runs a code value of \verb|e| and restore \verb|e|.
Escape, written \verb| .~ e | in MetaOCaml, expands a code value of \verb|e|.
Unlike run, escape is supposed to use in a fragment of code, that is, escape cannot be used as an alternative of run.
For example, the following MetaOCaml expression

@@

Cross-stage Persistence (CSP) is another primitive of multi-stage programming, which enable us to embed computed values into a code value.
This is an example of CSP.
\begin{verbatim}
        let plusone = fun x -> x + 1 in
        let y = plusone 2 in  .< y >.
\end{verbatim}
This program evaluates to \verb|.< 3 >.| not  \verb|.< plusone 2 >.|.
This is because the variable \verb|y| is introduced into a code fragment using CSP.
Therefore, the variable \verb|y| is embedded after it was calculated.
CSP is very important when we write practical programs 
because CSP enables us to use library functions in code fragments as in \verb|.<List.combine [1;2] ['a';'b']>.|.

@@

The main application of multi-stage calculi is program optimization.
A famous example is power functions.
Usual power functions take two arguments, which are the base and the exponent.
We can make a specialized power function for given exponents with multi-stage programming and 
optimize them by unrolling a loop in functions for given exponents.
The optimized power function for the exponent of \verb|3| looks like \verb|power3 = fun x -> x * x * x|.
\verb|power3| is faster than ordinary \verb|power| function because it contains no loop.
Multi-stage programming can optimize functions which take more than two arguments 
by generating a code fragment optimized to a given argument.

@@

Multi-stage programming can also optimize vector calculation.
For example, \verb|vadd| function, which takes two vectors and return the sum of them,
is implemented with a loop in many cases.
We can unroll \verb|vadd| function for a given vector length and optimize it with multi-stage programming.
\red{MetaOCamlで書いた長さ付きベクトルの例?}

@@

However, there is a sever problem in functions which are optimized by multi-stage programming.
Although unrolled \verb|vadd| function is optimized, we cannot use it for different vector length.
For example, when you optimize for the length of 5, you shouldn't use it for vectors of 3 lengths.
Otherwise, we will get a Segmentation Fault error.
This problem is serious but existing type systems for multi-stage calculi cannot prevent it.

@@

\subsection{Multi-stage Programming with Dependent Types}
\red{このsubsectionは短いので上とくっつけてもいいかもしれない}

@@

As we pointed out in the above section,
functions optimized with multi-stage programming can take only restricted values as arguments.
When you optimize \verb|vadd| function for the length of 5, you should it only for 5 length vectors.
We introduce dependent types into a multi-stage calculus
so that the type system can guarantee optimized functions are used properly.

In this paper, we design new multi-stage calculus \LMD by 
merging dependent types into existing multi-stage calculus \LTP\cite{Hanada2014}.
\LMD is multi-stage calculus which contains dependent types.
\LMD makes multi-stage programming more safer with the power of dependent types.

@@

We designed \LMD, a multi-stage calculus with dependent type.
\LMD is based on \LTP\cite{Hanada2014} by Hanada and Igarashi, which is a multi-stage calculus with CSP and
we introduced dependent types based on \LTP\cite{attapl}.
In this section, we check \LMD informally after checking \LTP and \LLF which are the basis of \LMD.

@@

\LTP\cite{Hanada2014} is a multi-stage calculus with CSP by Hanada and Igarashi.
In \LTP, brackets and escape are written $\TB_\alpha M$ and $\TBL_\alpha M$, respectively.
The type of $\TB_\alpha M$ is $\TW_\alpha \tau$ if $M$ has type $\tau$.
The type of $\TBL_\alpha M$ is $\tau$ if $M$ has $\TW_\alpha \tau$.
% Please notice that if $\TBL_\alpha M$ is well-typed, $M$ has a code type from the typing rule of \LTP.
In addition to normal $\beta$-reduction, there is a reduction rule for brackets and escape.
\begin{align*}
    \TBL_\alpha (\TB_\alpha M) \longrightarrow M
\end{align*}
It means escape cancels brackets.
% This reduction is called $\longrightarrow_\Lambda$ in \LMD.
% \red{この文はここに来るべきなのか? LTPの説明ではないが}

@@

Another important rule about a transition variable is 
that symbols with transition variables disappear 
when the empty transition $\epsilon$ is substituted to the transition variable.
The purpose of this rule is express \verb|run| without any special symbol.
In multi-stage calculus, \verb|run| is a very important operator which changes quoted code to the original code.
For example, $(\TB_\epsilon (\lambda x:\I.x))$ is equivalent to $(\lambda x:\I.x)$.
Therefore, $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))\ \epsilon$ becomes $\lambda x:\I.x$
In \LTP, \verb|run| is realized with application to the empty transition $\epsilon$.

@@

Second, we need kinding rules and well-formed kinding rules in order to extend \LMD with dependent types.
It was lucky that almost all rules are determined easily.
This is because multi-stage calculus and dependent types are almost orthogonal.
\red{orthogonal は抽象的すぎるか?}
Therefore, we can get kinding rules and well-formed kinding rules of \LMD just by 
attaching stage anotations to ones of \LLF.
For example, \KAbs-LF is a kinding rule for a dependent type in \LLF and \KAbs\ is a corresponding one.
\begin{center}
    \infrule{\G\vdash \tau :: * \andalso \G,x:\tau @A\vdash \sigma::J}{\G\vdash(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J)}{\KAbs-LF} \\[2mm]
    \infrule{\G\V \tau :: *@A \andalso \G,x:\tau @A\V \sigma::J@A}{\G\V(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J)@A}{\KAbs} \\[2mm]
\end{center}

@@

% 多段階計算のルール
The rules \TTB, \TTBL, \TGen, \TIns, and \TCsp\ are rules for a multi-stage calculus.
They are corresponding to quoting a code, unquoting a code, making a stage abstraction, 
application of $\epsilon$ stage to a stage abstraction, and cross-stage persistence, respectively.
Please check Hanada and Igarashi \cite{Hanada2014} and Tsukada and Igarashi \cite{Tsukada} for details of these rules.

@@

% \QPercentの説明
The rule \QPercent is something special because it has no background of syntax or reductions.
But we add this rule to \LMD\ because it is essential when you use dependent types in code type.
For example, we can utilize a dependently typed multi-stage calculus to generate code for matrix multiplication.
When we know sizes of matrices before the multiplication,
we can generate efficient code for multiplication by unrolling loops.

@@

\red{Besides,} multi-stage calculi also have a long history of research.
Davis revealed there is Curry-Howard correspondence between a multi-stage calculus and modal logic in $\lambda^\circ$\cite{davies1996temporal}.
However, $\lambda^\circ$ don't have operators for run and CSP.
\red{runはboldか? Besidesより適切な接続詞はないのか?}
Benaissa, et al. \cite{benaissa1999logical} study the relationship between multi-stage calculus and category theory or modal logic.
Taha and Sheard introduced run and CSP to a multi-stage calculus in \cite{MetaML}.
Additionally, Taha and Nielsen invented the concept of environment classifiers in $\lambda^\alpha$\cite{taha2003environment} and 
construct the type system for $\lambda^\alpha$, which can handle quoting, unquoting, run, and CSP.

In $\lambda^\TW$\cite{Tsukada}, Tsukada and Igarashi found modal logic which corresponds to a multi-stage calculus with environment classifiers and
show that run can be represented as application of $\epsilon$ to transition abstractions.
In \LTP\cite{Hanada2014}, Hanada and Igarashi extended $\lambda^\TW$ with CSP and discuss code residualization 
which allows us to dump the quoted code into an external file.

@@

Combination of multi-stage programming and another programming feature has been studied.
Oleg, Kameyama, and Sudo\cite{kiselyov2016refined} designed a type sysytem for multi-stage programming with mutable cells and
Oishi and Kameyama\cite{oishi2017staging} extended a multi-stage calculus so that it can handle control operators such as shift and reset.
